If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization

### 2025-08-15 GroupPolarizationStatmod/paper/intro-latest.tex
unexpectedly tweaking Intro

% \begin{itemize}
% \tightlist
% \item
We developed a generative model of opinion change to identify, if possible,
null opinion change models that could spuriously appear like group polarization
when naively measured using a Likert scale without accounting for the measurement procedure. 
We further applied our method to estimate the family-wise error rate of
detections of group polarization in a corpus of ten influential papers from 
across several decades and from various theoretical traditions. We turn this criticism of past
blunders constructive by providing a procedure to calculate what Cohen's $d$ should be
used to guarantee a family-wise error rate (i.e., Type I error rate) of 5\% or less.

@@

\subsection{Group polarization research}

@@

@@

\subsection{How study design can induce false detections of group polarization}
\label{ssubsec:false-detections}

>>>>

% \begin{itemize}
% \tightlist
% \item

We developed a generative model of opinion change to identify, if possible,
\emph{simple consensus} latent distributions that could spuriously appear like group polarization.
This type of problem is known to happen when ordinal measurements are taken of continuous data
without statistically accounting for the measurement procedure itself~\cite{Liddell2018}.
We further applied our method to estimate a best-case probability that 
a published detection of group polarization is false, across 57 experimental
conditions in a corpus of ten journal articles curated for their prominence and
representativeness of group polarization research. 
We turn this criticism of past blunders constructive by providing a procedure to calculate what Cohen's $d$ should be
used to guarantee a family-wise error rate (i.e., Type I error rate) of 5\% or less.

@@

\subsection{Group polarization}

@@

- Group polarization is a type of \emph{consensus}. \emph\emph{Simple consesus} is consensus where
\emph{only} the opinion variance of a group changes, not the mean opinion of the group.

\subsubsection{Group polarization research}

@@

\subsubsection{Failure to account for the measurement process can induce false detections}
\label{ssubsec:false-detections}

### 2025-08-15 GroupPolarizationStatmod/paper/ms.tex
unexpectedly tweaking Intro

\emph{Spurious} group polarization is when mean observed opinions seem to increase in 
extremity when the latent mean is held constant. Call the collection of
latent null model distribution parameters $\lambda = (\mu, \sigma_{pre}, \sigma_{post})$.
The group polarization opinion shift can then be written as a function of $\lambda$,
$g(\lambda)$. Formally, then, spurious group polarization is plausible if 
there exists at least one $\lambda$ for which $g(\lambda) = g_{pub} > 0$, where
$g_{pub}$ is the published group polarization shift value.

@@

\subsection{Web app controls algorithm to find parameters inducing spurious group polarization}

To calculate the false discovery rate for each experimental condition, we first must
identify a constant latent mean, $\mu$, and latent variances $\sigmapre$ and
$\sigmapost$ that generate observation distributions\mt{define earlier?} with
reported mean pre- and post-deliberation opinions and resulting shift $s$ (calculated
using Equation~\ref{eq:group-polarization}). Define the collection of these
parameters $\lambda = \{\mu, \sigmapre, \sigmapost\}$. We identify $\lambda$ through a
hillclimbing optimization algorithm that finds $\lambda$ that minimizes the difference
between reported pre- and post-deliberation mean opinions and simulated ones
calculated with Equation~\ref{eq:meanobs}.

@@

% We first identified latent opinion
% parameters that yield spurious group polarization in an asymptotic $N\to\infty$
% model. This gives us our first suggestion that experimental conditions may be
% plausibly false. Briefly, we identified latent opinion distribution parameters
% that generated spurious group polarization in 54 out of 57 experimental
% conditions. In other words, 95\% of all group polarization detections are
% plausibly spurious due to ceiling artifacts from the continuous latent
% opinions-ordinal measurement experimental design. We then analyze the family-wise
% error ($\alpha$) and false detection rate (FDR) 
% for each experimental condition and aggregated 
% up to the study level and across all studies and experimental conditions. 
% We estimate the lower twentieth percentile of experimental conditions 
% have \emph{at least} $\alpha \leq 0.16$, 
% with median $\alpha \geq 0.27$ and upper eightieth percentile with $\alpha \geq
% 0.40$ (Table~\ref{tab:quantiles}\mt{A.  –but don't have A and B; TODO: separate
% FWER and FDR tables, only use 0, 20, 50, 80, 100 for probs. Rename 0 -> min, 100 ->
% max.}). For the fiftieth and eightieth percentiles this
% corresponds to false discovery rates of 0.75 and 0.82, respectively. 
% We emphasize that these estimates and others detailed below are lower bounds on
% $\alpha$ and FDR since we are using what is likely an overgeneralized,
% underspecified statistical model that does not account for all the potential
% variance in experimental outcomes, which would increase error rates.
% We close our Analysis by calculating which $d^*$ limits family-wise error rates to
% $\alpha \leq 0.05$. This is greater than $d^* = 0.8$, classically taken to 
% represent ``high'' significance, in all but six experimental conditions.

@@

We used our formal distribution model of group polarization experiments
to identify plausible latent simple consensus distributions that generate spurious
group polarization in observed ordinal distributions 
in the asymptotic case of taking $N \to \infty$ ordinal measurements.
Across the ten studies we analyzed, we excluded 3 out of 60 experimental
conditions that did not claim to be positive findings of group polarization.
Of the remaining 57, we identified simple consensus $\lambda$ that generate spurious
group polarization in 54 experimental conditions. In the other three, we did
identify $\lambda$ that generate spurious group polarization, but the initial
simulated distributions were highly polarized. We rejected $\lambda$ generated for
two spurious group polarization from Schkade, et al., (2010) because
the histogram plots of the observed opinion distributions are not
polarized~\cite[Figure 1, p. 234]{Schkade2010}. The other condition where we
found but rejected spurious group polarization $\lambda$ was in a question 
about a ``bad teacher'' in Myers (1975)~\cite{Myers1975a}, where lower values
signify increasingly worse impressions of the teacher and higher values signify
increasingly good impressions. It is unlikely that a
``bad teacher'' would be rated as good in this case, so we excluded this 
from our analysis as well. This exclusion may be overly generous since Myers (1975)
did not provide details beyond mean values and opinion shifts, and the article was
published before open data practices.

>>>>

\emph{Spurious} group polarization is when mean observed opinions seem to increase in 
extremity when the latent mean is held constant. Call the collection of
latent null model distribution parameters $\lambda = (\mu, \sigma_{pre}, \sigma_{post})$.
The group polarization opinion shift can then be written as a function of $\lambda$,
$g(\lambda)$. Formally, then, spurious group polarization is plausible if 
there exists at least one $\lambda$ for which $g(\lambda) = g_{pub} > 0$, where
$g_{pub}$ is the published group polarization shift value.
This could be thought of as sampling the observed ordinal distributions 
$N \to \infty$, collecting ordinal measurements of samples from 
each latent distribution corresponding to each phase, pre- and post-deliberation.

@@

\subsection{Interactive algorithm to find parameters inducing spurious group polarization}

To identify plausibly false detections of group polarization, we searched numerically
for latent parameters $\lambda = (\mu, \sigmapre, and \sigmapost)$ that generate
spurious group polarization, as explained immediately above. We developed an *ad hoc*
hillclimbing-style algorithm that finds $\lambda$ that finds $\lambda$ by minimizing
the difference between published group polarization shifts $g_pub$ and calculated shifts,
$g_\lambda$.
calculated with Equation~\ref{eq:meanobs}.

@@

@@

% We used our formal distribution model of group polarization experiments
% to identify plausible latent simple consensus distributions that generate spurious
% group polarization in 
Of the remaining 57, we identified simple consensus $\lambda$ that generate spurious
group polarization in 54 experimental conditions. In the other three, we did
identify $\lambda$ that generate spurious group polarization, but the initial
simulated distributions were highly polarized. 
We rejected $\lambda$ generated for
two spurious group polarization from Schkade, et al., (2010) because
the histogram plots of the observed opinion distributions are not
polarized~\cite[Figure 1, p. 234]{Schkade2010}. The other condition where we
found but rejected spurious group polarization $\lambda$ was in a question 
about a ``bad teacher'' in Myers (1975)~\cite{Myers1975a}, where lower values
signify increasingly worse impressions of the teacher and higher values signify
increasingly good impressions. It is unlikely that a
``bad teacher'' would be rated as good in this case, so we excluded this 
from our analysis as well. This exclusion may be overly generous since Myers (1975)
did not provide details beyond mean values and opinion shifts, and the article was
published before open data practices. Our corpus had a total of 60 experimental conditions, but we excluded 3 out of 60 
that did not claim to be positive findings of group polarization.

- Address sigmas, note differences

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{Figures/3.1/latent-sd-cdf.png}
    \caption{\textbf{}}
  \label{fig:}
\end{figure}
